
# Generate interactive ROC plots from R using ggplot

Most ROC curve plots obscure the cutoff values and inhibit
    interpretation and comparison of multiple curves. This attempts to address
    those shortcomings by providing plotting and interactive tools. Functions
    are provided to generate an interactive ROC curve plot for web use, and
    print versions. An interacive Shiny application is also included.
    
## Installation

This package is not currently on CRAN. To install, use `devtools`

```{r install}
devtools::install_github("sachsmc/plotROC")
```

## Basic usage

See examples live [here](http://sachsmc.github.io/plotROC)

```{r test-a, fig.keep='none', results = 'asis', echo = TRUE, fig.width=6, fig.height=6}
library(plotROC)
D.ex <- rbinom(100, 1, .5)

rocdata <- calculate_roc(rnorm(100, mean = D.ex), D.ex)

myrocplot <- ggroc(rocdata)

cat(svgize_roc(myrocplot, cutoffs = rocdata$c, font.size = "12px", prefix = "a"))
```

That is an svg element. Try zooming in!

## Generate an interactive plot to view in Rstudio viewer or web browser

```{r inter, eval = FALSE}
standalone_svg_roc(rocdata)
```


# Themes and annotations

```{r test-b, fig.keep='none', results = 'asis', echo = TRUE, fig.width=6, fig.height=6}
library(ggthemes)
myplot2 <- myrocplot + theme_igray()

cat(svgize_roc(myplot2, cutoffs = rocdata$c, font.size = "12px", prefix = "b"))
```

```{r test-c, fig.keep='none', results = 'asis', echo = TRUE, fig.width=6, fig.height=6}

myplot2 <- myrocplot + theme_solarized() + annotate("text", x = .5, y = .5, label = "Reference line", hjust = 0, vjust = 1)

cat(svgize_roc(myplot2, cutoffs = rocdata$c, font.size = "12px", prefix = "c"))
```

